---
title: "Beta-convergence and Sigma-divergence - a convergence paradox?"
author: "Rok Hrzic"
date: "30 January 2019"
output: html_document
---

## Motivation

In health convergence literature, we often come across a result of simultaneous beta-convergence and sigma-divergence. The first one says that areas with initially poor performance have been improving faster than those with initially good performance. The second says that the overall dispersion is increasing in time. It is not entirely clear whether this occurance is a fallacy (regression to the mean) or a real occurence in health demography. 

The aim is to measure (unadjusted) beta- and sigma-convergence on a variety of simulated datasets to build understanding and intuition on this topic.

## Loading packages

```{r}
require(dplyr)
require(purrr)
require(ineq)
require(ggplot2)
require(gridExtra)
```

## Datasets and their analysis

I will generate datasets that reflect various scenarios. They will include 10 fictional regions with different developments of life expectancy over 20 years.

Each of the scenarios will be evaluated for beta-convergence via a regression of annualised growth of LE on the initial value of LE, and for sigma-convergence via a calculation of the Gini coefficient, the absolute gap, the coefficient of variation and Thiel index. 

```{r}

analyse_scenario <- function(scenario) {
  
  scenario <- scenario %>% mutate(area = as.factor(area)) %>%
    group_by(year) %>% mutate(aveLE = mean(LE, na.rm = T)) %>% ungroup
  
  a <- ggplot(scenario) +
    geom_line(aes(x = year, y = LE, group = area, colour = area)) +
    geom_point(aes(x = year, y = aveLE))
  
  b <- summary(lm((LE/LE0)/20 ~ LE0, scenario))$coefficients[2,] %>% format(x, scientific = F)

  c <- ggplot(scenario) +
    geom_point(aes(x = LE0, y = LE, colour = area))
  
  d <- scenario %>% group_by(year) %>% 
    summarise(Gini = ineq(LE, type = "Gini"),
              Range = max(LE, na.rm = T) - min(LE, na.rm = T),
              COV = ineq(LE, type = "var"),
              Variance = var(LE, na.rm = T))
  
  e <- ggplot(d, aes(x = year, y = Gini)) +
        geom_point() +
        geom_smooth(method = "lm")
  
  f <- ggplot(d, aes(x = year, y = Range)) +
     geom_point()+
        geom_smooth(method = "lm")
  
  g <- ggplot(d, aes(x = year, y = COV)) +
     geom_point()+
        geom_smooth(method = "lm")
  
  h <- ggplot(d, aes(x = year, y = Variance)) +
     geom_point()+
        geom_smooth(method = "lm")
  
  i <- grid.arrange(e,f,g,h, nrow = 2)
  
  j <- summary(lm(Gini ~ year, d))$coefficients[2,] %>% format(x, scientific = F)


  results <- list(a, b, c, i, j)
  
  return(results)
}
```

## Static scenarios - no real changes in rank

### Scenario 1: Stability

In this scenario, there will be no change in life expectancy. I expect parallel flat lines on plots, no beta-convergence and no sigma-convergence. 

```{r}
scenario1 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  group_by(year, area) %>%
  mutate(LE = LE0+rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

analyse_scenario(scenario1)
```

Indeed, there was no statistically significant beta-convergence and all sigma-convergence measures are flat. 

### Scenario 2: We improve together

Now we add a linear trend, but keep the lines parallel (no change in rank or gap).

```{r}

scenario2 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  group_by(year, area) %>%
  mutate(LE = LE0 + rnorm(1,0,0.1) + 2/20 * year) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup


analyse_scenario(scenario2)

```

We find no statistically significant beta-convergence, but (on some runs) significant sima-convergence when considering Gini and CoV - not when considering range and variance.

The literature (e.g. Sala-i-Martin, 1996; Jannsen, 2017) says that beta-convergence is a necessary precondition of sigma-convergence, but I think this is true only when using measures of dispersion such as range, sd and variance, which do not depend on the mean. In fact, if we go back to the Sala-i-Martin publication that summarises the classical approach to convergence analysis, they equate dispersion with **distance** - e.g. gap or range. In this case, indeed, it would be impossible to have a narrower gap between units at the end of observation without different growth rates between units with the initially poorer performing units enjoying higher rates than the initially better performing units.

However, relative measures of dispersion, such as Gini, CoV or Theil, do not measure the gap or range per-se, but rather measure it's size relative to the mean. So if the central tendency of the distribution increases with the range between units staying constant, we can indeed expect a significant decrease in these measures of dispersion. In this case, however, they measure the extent of relative inequality; see also Mackenbach's (2015) discussion of absolute and relative mortality inequalities.

## Unequal growth scenarios without rank change

Here I include scenarios where the good get better and/or the poor get worse, i.e. the Matthew effect.

### Scenario 3: Good get better

Here the top regions improve life expectancy at twice the rate as other regions.

```{r}

scenario3 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = ifelse(area %in% c(8,9,10), LE0 + year * 2/20, LE0 + year * 1/20)) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

analyse_scenario(scenario3)

```

We get significant beta-divergence (beta is positive!) as well as sigma-divergence, which is to be expected in such a sitiation.

### Scenario 4: The worse-off stagnate

Here most regions will improve at a certain pace, with the worst regions experiencing a much more languid pace of development. 

```{r}

scenario4 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = ifelse(area %in% c(1,2,3), LE0 + year * 1/20, LE0 + year * 2/20)) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

analyse_scenario(scenario4)

```

As expected, beta-divergence and sigma-divergence are the results. 

### Scenario 5: Regression to the mean

Finally, here we test a scenario that is supposed to be one of the causes of the beta-convergence, sigma-divergence paradox. We will pretend that the worst of the regions starts with either one or two significant negative outliers, but then recovers the pace of development with the others.

```{r}
scenario5 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = LE0 + year * 2/20) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

scenario5[scenario5$year == 1 & scenario5$area == 1,]$LE <- scenario5[scenario5$year == 1 & scenario5$area == 1,]$LE-2

analyse_scenario(scenario5)
```

We fail to create a significant beta-convergence here, but get significant sigma-convergence - the opposite of what we are looking for. Let us try two outliers. 

```{r}
scenario5[scenario5$year == 2 & scenario5$area == 1,]$LE <- scenario5[scenario5$year == 2 & scenario5$area == 1,]$LE-1.5

analyse_scenario(scenario5)
```

Same situation. It seems to me that the outlier hypothesis is not sufficient, unless it impact a large number of areas and causes rank changes (see later).

### Scenario 6: Polarisation

In this scenario, two convergence clubs form.

```{r}
scenario6 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = ifelse(area %in% c(1,2,3,4,5), LE0 - year * 1/20 * (area-1)/5, LE0 + year * 1/20 * (10-area)/5)) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup
analyse_scenario(scenario6)
```

We can see that the range indicator is very insensitive to changes in the middle of the distribution, while the other indicators successfuly show an increase in inequality - sigma-divergence. As expected, we can also confirm beta-divergence.

## Rank changes

In this set of scenarios, we will get areas to change rank.

### Scenario 7: The laggards catch up

The laggards here have four-times the LE growth compared to the rest.

```{r}

scenario7<- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = ifelse(area %in% c(1,2,3), LE0 + year * 4/20, LE0 + year * 1/20)) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

analyse_scenario(scenario7)

```

This looks like a classic catching up scenario, and it results in both beta- and sigma-convergence.

### Scenario 8: The laggards catch up, but the middle flounders

In this scenario, the top performers keep improving, the bottom performers catch up, and the middle sort of stays still.

```{r}

scenario8 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = ifelse(area %in% c(1,2,3), LE0 + year * 3/20, LE0 + year * 1/20)) %>%
  mutate(LE = ifelse(area %in% c(5,6,7), LE0 - year * 0.5/20, LE)) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

analyse_scenario(scenario8)

```

This results in both beta- and sigma-convergence, so no luck with recreating the paradox yet. 

### Scenario 9: The laggards catch up, but the middle deteriorates

In this scenario,the bottom performers catch up, and the middle performers actually loose ground.

```{r}

scenario9 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = ifelse(area %in% c(1,2,3), LE0 + year * 2/20, LE0 + year * 1/20)) %>%
  mutate(LE = ifelse(area %in% c(5,6,7), LE0 - year * 2/20, LE)) %>%
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>% 
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup

analyse_scenario(scenario9)

```

This configuration recreates the paradox. We have the very bottom performers doing very well (catching up), and the mid-tier performers losing **substantial** ground. It is very interesting that the scenario is very dependent on how many areas we include in which group, their initial conditions, and how much ground the middle is losing. 

The next step is to hash out these limits more precisely by breaking apart the formulae for beta- and sigma-convergence and using simulations to detect critial points in each.

### Scenario 10: New entrants
In this scenarios, new units enter at a certain time (in our case, 5 years through the observation). In line with EU expansion, they will enter at the bottom of the distribution.

```{r}

scenario10 <- expand.grid(area=1:10, year=1:20) %>% 
  mutate(LE0 = as.numeric(rep(sort(rnorm(10, 80, 2)), 20))) %>%
  mutate(LE = LE0 + 2/20 * year) %>%
  mutate(LE = ifelse(area %in% c(1,2,3) & year < 6, NA, LE)) %>%
  mutate(LE0 = ifelse(area %in% c(1,2,3), LE0 + 2/20 * 6, LE0)) %>% 
  group_by(year, area) %>%
  mutate(LE = LE + rnorm(1,0,0.1)) %>%
  ungroup %>%
  group_by(area) %>% 
  mutate(LE0 = first(na.omit(LE))) %>%
  ungroup


analyse_scenario(scenario10)

```

The new entrants scenario results in sigma-divergence (false due to uncorrected step).

## Taking apart beta- and sigma-convergence formulae

Unadjusted beta-convergence occurs when the coefficient in the following formula is statistically significant negative:

(yT/yt0) / (T - to) = a + b * yt0

Parsing this into words, the outcome is the annualised growth rate and the only predictor is the outcome at the start of observation.

I will focus on the Gini coefficient here, since the above simulations show that the other relative dispersion indicators exhibit the same qualitative behaviour.

An empirical Gini formula is given by Cowell as

G = 2 / (n^2 * yhat) * [y(1) + 2* y(2) + 3 * y(3) ... n*y(n)] - (n + 1)/n